Catastrophic Goodhart: regularizing RLHF with KL divergence does not mitigate heavy-tailed reward misspecification

### 2024-10-24 paper/main.tex
add icml-mi version of paper

@@

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2024

% ready for submission
\usepackage{neurips_2024}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2024}

% to compile a camera-ready version, add the [final] option, e.g.:
%     \usepackage[final]{neurips_2024}

% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2024}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{graphicx}       % figures
\usepackage{enumitem}
\usepackage{subfiles}       % For long proof of Theorem 5

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

\title{Catastrophic Goodhart: regularizing RLHF with KL divergence does not mitigate heavy-tailed reward misspecification}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{%
  Thomas Kwa
  Independent / FAR Labs
  \texttt{kwathomas0@gmail.com} \\
  \And
  Adrià Garriga-Alonso \\
  FAR AI \\
  \texttt{adria@far.ai} \\
}

\begin{document}

\maketitle

@@

Kullback-Leibler (KL) divergence constraints in reinforcement learning (RL) are employed to stay in regimes where the objective is accurate enough.
Some on-policy \citep{trpo,ppo} and many off-policy \citep{abdolmaleki18mpo,jaques2019way} policy gradient algorithms employ KL constraints or penalties during optimization to prevent the policy from deviating too much from the data collection distribution. This ensures that estimates of each action's advantage are good enough to update the policy in a helpful way.

@@

In RLHF, it is common to use the regularization term $\beta D_{KL}\left(\pi \| \pi_0 \right)$ to prevent the learned policy from deviating too much from the base policy, which can prevent unstable behavior or overfitting to the reward model. If our reward model gives reward $U$, then the optimal policy for RLHF with a KL penalty is \[
\arg \max_{\pi} \mathbb{E} [U(\pi)]-\beta D_{KL}\left(\pi \| \pi_0 \right).
\]

@@

A distribution $P$ over $\mathbb R$ with cumulative distribution function (CDF) $F_P$ is heavy-tailed if its tail function $\bar F_P(x) \triangleq 1 - F_P(x)$ satisfies
\[
\lim _{x \rightarrow \infty} e^{t x} \bar{F}(x)=\infty \quad \text { for all } t>0
\]
Heavy-tailed distributions are well-known in statistics to have a higher probability of producing a single extreme value. For example, if the sum of two independent variables from heavy-tailed distributions is large, it is most likely due to one extreme sample rather than two equally large samples. \citep{wierman2013catastrophes}

@@

Reward misspecification is also studied by \citep{lambert2024alignment}, \citep{laidlaw2024preventing}, and others. Laidlaw et al show that a KL penalty between action distributions can be ineffective, and propose instead KL-regularizing state occupancy measure. Our results show an inherent weakness of KL divergence, including when applied to state occupancy measure.

@@

We prove that in many cases, $V \to 0$ in the limit of optimization for some proxy $U$. We call this phenomenon ``catastrophic Goodhart'', after Goodhart's law: ``when a measure becomes a target, it ceases to be a good measure'' \citep{Strathern_1997}. In these cases, the end result of optimizing for a proxy of $V$ is no better than not optimizing at all. However, in other cases, $V \to \infty$ despite some reward misspecification; in these cases the reward misspecification is not severe enough to prevent good outcomes.

@@

Theorems~\ref{thm5} and \ref{thm6} (Section~\ref{sec:theoretical-conditioning} in the appendix) show that the relationship of catastrophic Goodhart to heavy-tailed error is not just a quirk of KL divergence by using a different model of optimization based on conditioning on high reward values. Under this model (and given additional regularity conditions), it is also true that heavy-tailed error results in catastrophic Goodhart, and light-tailed error plus independence results in arbitrarily large utility. All proofs are in the appendix.

\subsection{Optimizing the policy with a KL term\label{sec:theoretical-kl}}

\subsubsection{Heavy-tailed distributions}
\begin{theorem}
\label{thm1}
Given any heavy-tailed reference distribution
\(Q\) over \(\mathbb R\) with mean \(\mu_Q\), and any
\(M, \epsilon > 0\), there is a distribution \(P\) with mean \(\mu_P>M\)
and \(D_{KL}(P \| Q) < \epsilon\).
\end{theorem}

@@

\hypertarget{theorem-about-RLHF-with-KL-penalty}{%
\subsubsection{RLHF with KL penalty under heavy-tailed return distribution}}

We now adapt our result to the case where our policy is a language model and we are training it using RLHF. We are now applying KL divergence over the policies, which are a different distribution than the returns $U$. But first we need to formally define the properties of RLHF on language models which cause the result to hold, namely that when considered as a Markov decision process (MDP), environmental transitions are deterministic and return depends only on the final state reached.

@@

\begin{itemize}
\item
  The transition function
  \(P: \mathcal{S} \times \mathcal{A} \to \mathcal{S}\) is
  deterministic, i.e., for each state \(s \in \mathcal{S}\) and action
  \(a \in \mathcal{A}\), there exists a unique state
  \(s' \in \mathcal{S}\) such that \(P(s' | s, a) = 1\).
\item
  There is a set of sink states \(E \subseteq \mathcal S\) that
  terminate a trajectory, which is disjoint with the set of start
  states.
\item
  Returns are Markovian; that is, for any two trajectories
  \(\tau=(s_1, a_1, \dots, s_n), \tau'=(s_1', a_1', \dots, s_n'),\) if
  \(s_{n}= s'_{n}\), then \(\tau\) and \(\tau'\) have identical return
  distributions. Equivalently, for the trajectory random variable
  \(T=(S_1, A_1, \dots)\) distributed according to any policy, with
  return \(G\), \(G \bot\!\!\!\!\!\ \bot (S_{<i}, A_{<i}) \ |\ S_i\) for
  any \(i \ge 1\).
\end{itemize}

Note: If we consider the state to be the sequence of tokens output so far, an action to be the next token sampled, and reward to be the reward model's output (as usual), sampling from a language model and applying RLHF is well-modeled
as a DMRMDP, since the state deterministically results from the last token and previous state, and returns depend only on the final state.

\begin{theorem}
\label{thm2}
Let \(W = (\mathcal S, \mathcal A, P, R)\) be
a deterministic-transition MDP with Markovian returns. Given \(W\) we
define the function that takes policies to trajectories
\(Tr: (S \to \Delta A) \to \Delta(S \times A)^*\), and the average
return function \(g: (S \times A)^* \to \mathbb R\) which induces a
function \(G: \Delta(S \times A)^* \to \Delta \mathbb R\). Let
\(\pi_0: \mathcal S \to \Delta \mathcal A\) be some reference policy. If
\(G \circ Tr(\pi_0)\) is heavy-tailed with finite mean \(\mu_Q\), then
for any \(M, \epsilon > 0\), there is a policy \(\pi\) with mean return
\(\mathbb E[U | U \sim G \circ Tr(\pi)] > M\) and
\(\mathbb E_{s \in T, T\sim Tr(\pi)}[D_{KL}(\pi(s) \| \pi_0(s))] < \epsilon\).
\end{theorem}

\subsubsection{If V is light-tailed, \texorpdfstring{$\mathbb E[V] \to 0$}{} as \texorpdfstring{$D_{KL} \to 0$}{}}

\begin{theorem}
\label{thm3} If \(V\) is light-tailed, \(\mathbb E_Q[V]\)
is zero, and \(d = D_{KL}(P \| Q)\) is bounded, then
\(\mathbb E_P[V]\) is bounded, and \(\mathbb E_P[V] \to 0\) as
\(d \to 0\).
\end{theorem}

\newtheorem{corr}{Corollary}
\begin{corr}
\label{cor:heavy-tailed}
    Theorems~\ref{thm2} and \ref{thm3} imply that when the reward has heavy-tailed error and a policy is regularized severely enough to have KL divergence values approaching zero, reward can go to infinity while utility goes to zero.
\end{corr}

\hypertarget{light-tails-independence-imply-mathbb-ev-to-infty}{%
\subsubsection{\texorpdfstring{Light tails + independence imply
\(\mathbb EV \to \infty\)}{Light tails + independence imply \textbackslash mathbb EV \textbackslash to \textbackslash infty}}\label{light-tails-independence-imply-mathbb-ev-to-infty}}

@@

\subsection{Conditioning as alternate model of optimization\label{sec:theoretical-conditioning}}

Although we think a KL divergence penalty or cap is the most realistic setting for RLHF, it is not the only model of optimization where heavy-tailedness of the error determines whether catastrophic Goodhart occurs. Consider another model of optimization where $U = X+V$ as before, but we simply condition on $U$ being higher than some threshold $t$.\footnote{This could model a satisficing agent that takes random acceptable actions.} Then we are interested in the quantity $\lim_{t \to \infty} \mathbb E[V | X + V \ge t]$. If we slightly strengthen the heavy-tailedness and light-tailedness assumptions, heavy-tailed error results in catastrophic Goodhart, while light-tailed error results in arbitrarily high expected utility.

\subsubsection{Conditioning with heavy-tailed error produces catastrophic Goodhart}
\begin{theorem}
    \label{thm5} Let $X$ and $V$ be two independent random variables with CDFs $F_X$ and $F_V$ and tail functions $\bar F_V \triangleq 1 - F_V$, $\bar F_X \triangleq 1 - F_X$ such that
    \begin{itemize}
        \item $V$ has a finite mean.
        \item $X$ is subexponential; that is, $\lim_{x\to\infty}\frac{\text{Pr}(X_1+X_2>x)}{\text{Pr}(X>x)} = 2$ if $X_1, X_2$ are two independent samples from $X$. This is a slightly stronger property than being heavy-tailed.
        \item The tail of $V$ is sufficiently lighter than the tail of $X$ that \(\lim_{t\to\infty}\frac{t^p\bar F_V(t)}{\bar F_X(t)} = 0\) for some \(p > 1\).
    \end{itemize}
    Then $\lim_{t \to \infty} \mathbb E[V | X + V \ge t] =\mathbb{E}[V]$; that is, catastrophic Goodhart occurs in the limit of optimization for $U=X+V$.
\end{theorem}

The proof is particularly long and, in our opinion, unenlightening, and will be included in the supplementary material. It requires expressing the conditional expectation in question as $\frac{\int_{-\infty}^\infty vf_V(v)\text{Pr}(X>t-v)} {\int_{-\infty}^\infty f_V(v)\text{Pr}(X>t-v)}$, then partitioning the interval $(-\infty, \infty)$ into four regions and bounding the integrand in the numerator above by a different quantity in each region.

\subsubsection{Conditioning with light-tailed error produces arbitrarily high utility}
\begin{theorem}
    \label{thm6}
    Let $X, V$ be independent random variables such that $\lim_{t\to\infty}\frac{\bar{F}_X(t+1)}{\bar{F}_X(t)}=0$. (This implies that X has tails that are dominated by $e^{-cx}$ for any c, though it's a slightly stronger claim because it requires that X not have large jumps in the decay of its tails.)
    Then for any V with a finite mean which has no upper bound, $\lim_{t\to\infty}\mathbb{E}[V|X+V > t] = \infty$. 
\end{theorem}

Theorem 6 generalizes a consequence of the "Regressional Goodhart Identity" in \citep{gao2023scaling}.

@@

To empirically test whether reward is heavy-tailed, we consider two lines of evidence: examining the distributions directly through random sampling, and finding adversarial token sequences that get high reward. We examine one small and one medium reward model that performed reasonably well on RewardBench \citep{lambert2023rewardbench}. The small model is an OpenAssistant model based on Pythia 1.4B, and the medium model is Starling 7B-alpha \citep{starling2023}\footnote{https://huggingface.co/berkeley-nest/Starling-RM-7B-alpha}.

Because random sampling is inefficient at probing the tail past $10^4$ or so samples, we also optimize Starling 7B-alpha for reward in an effort to find token sequences that are extreme positive outliers in terms of reward. We considered Greedy Coordinate Gradient (GCG) from \citep{zou2023universal}, a method used to find adversarial suffixes that circumvent jailbreaking, but decided on a modification to GCG called Accelerated Coordinate Gradient (ACG) from the blog post \citep{haizelabs2024acg}, which the authors claim is 38x faster.

Hyperparameters for ACG are given in Table 1 and were chosen to maximize reward and speed.

\begin{table}
    \centering
    \begin{tabular}{|c|c|}
        \hline
        Parameter & Value \\
        \hline
        Context length & 133 \\
        \hline
        Iterations & 1000 \\
        \hline
        Candidates per seq. position (k) & 3 \\
        \hline
        Annealing starting value & 9 \\
        \hline
        Annealing ending value & 2 \\
        \hline
    \end{tabular}
\caption{Hyperparameters for ACG}
\end{table}

@@

When sampling token sequences, both the Pythia model on random inputs (Figure \ref{fig:pythia-random}) and Starling 7B-alpha on Llama-generated inputs (Figure \ref{fig:starling-llama}) appear to be approximately normal and therefore light-tailed. Starling on random inputs is ambiguous, with the exponential Q-Q plot having an outlier that could indicate heavy-tails, but the Hill estimator is consistent with light tails. Because Llama-7B-chat is a more reasonable base model than a completely random policy, we think that Starling 7B-alpha is more likely to be light-tailed for the purposes of our theoretical results.

\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth]{images/reward_plots_random_starling_30k_se.png}
    \caption{Plots of the distribution of reward from 30000 random length-1024 token sequences to Starling 7B-alpha. Clockwise from top left: The histogram shows a unimodal distribution with a slight right skew. The normal probability plot indicates the data are heavier-tailed than normal. The Hill estimator (error bars are standard error) appears to be 0.20 for higher values but fluctuates for lower values. The exponential probability plot of the right half of the distribution is consistent with either light or heavy tails (under heavy tails the slope would go to infinity).}
    \label{fig:starling-random}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{images/reward_plots_llama_starling_16k.png}
    \caption{Plots of the reward distribution from 16000 token sequences generated by Llama 7B-chat of length $\le 133$, starting with five random tokens. Clockwise from top left: A histogram shows the reward distribution has a left skew. Normal probability plot suggests reward is approximately normal and thus light-tailed. The Hill estimator plot should stabilize if the distribution is heavy-tailed, but it does not; thus there is no evidence the distribution is heavy-tailed. The exponential probability plot also indicates light tails, because the curve is bending downwards. }
    \label{fig:starling-llama}
\end{figure}

The ACG results need some interpretation. The KL divergence between two distributions $P$ and $Q$ if $P$ is the same as $Q$ a fraction $1-f$ of the time, but is some value $x$ a fraction $f$ of the time is given by

\[ D_{KL}(P \| Q) = \left[(1-f)q(x) + f\right] \log \left(\frac{(1-f)q(x) + f}{q(x)}\right) + (1-f) \log (1-f) (1 - q(x)) \]

When f is small but much larger than $q(x)$, we approximate this to first order as $D_{KL}(P \| Q) \approx f \log \left(\frac{f}{q(x)}\right)$. In Theorems 1 and 2, we prove that when error is sufficiently heavy-tailed, a policy that gets extremely large reward a small fraction of the time will achieve high expected reward with low KL divergence. This is not the case here because the rewards achieved through ACG were small and the log-probabilities extremely negative. For example, a policy that matches Llama 2-chat's base reward 99\% of the time and uses the best input generated by ACG 1\% of the time will have KL divergence from Llama 2-chat of $(\log(0.01) - 1339.70) = 13.35$ nats, but reward only about $6.02 * 10^{-3}$ greater than the base model, far less than can be obtained with the same reward by conditioning.

@@

The low-KL policies that result in Catastrophic Goodhart is not a unique optimal policy, just one family of high-performing policies. When optimizing $\mathbb{E} [U(\pi)]-\beta D_{K L}\left(\pi, \pi_0 \right)$, the outcome depends on RL training dynamics; it could be that $D_{KL} \to 0$ causing catastrophic Goodhart, but more likely both terms will go to infinity, potentially allowing $V \to \infty$.

@@

Theorems 2 and 3 do not require any independence assumption, but theorems 4, 5, and 6 requires that error $X$ and utility $V$ are independent, which seems to be violated in practice. Future work could weaken this assumption, although intuitively obvious ways to weaken it result in the statement being false. \footnote{Suppose that error $X$ is light-tailed conditional on any value of $V$, but our proxy is merely unbiased ($\mathbb E[X|V=v]=0$ for all $v$). Then the limit of $V$ under optimization for $X+V$ still depends on the relationship between $X$ and $V$. If they are independent, theorem 6 says that \(\lim_{t \to\infty} \mathbb E[V | X + V \ge t] = \infty\). But if \(V \sim N(0, 1)\), and \(X | V \sim N(0, 4)\) when \(V \in [-1, 1]\), otherwise \(X=0\), then \(\lim_{t \to\infty} \mathbb E[V | X + V \ge t] = 0\).}

@@

We did not search the entire space of token sequences, we cannot rule out that reward is heavy-tailed enough to cause catastrophic Goodhart in some situations. While it is intractable to search the more than $10^{2000}$ possible token sequences, future work could get more evidence through more powerful optimization methods or MCMC.

@@

We have argued that the purpose of the KL divergence regularization in RLHF is to mitigate reward misspecification. However, we have also proven that, when errors in the reward function are heavy-tailed, it cannot serve this purpose: even with zero KL divergence, there are policies that achieve very high misspecified reward and no actual reward.

>>>>

%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

@@

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2024} with \usepackage[nohyperref]{icml2024} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage[accepted]{icml2024}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2024}

% For theorems and such
\usepackage{mathtools}

% Packages copied from neurips
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{graphicx}       % figures
\usepackage{enumitem}
\usepackage{subfiles}       % For long proof of Theorem 5

% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
% \icmltitlerunning{Catastrophic Goodhart}
\begin{document}

\twocolumn[
\icmltitle{Catastrophic Goodhart: regularizing RLHF with KL divergence does not mitigate heavy-tailed reward misspecification}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2024
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Thomas Kwa}{indp,farlabs}
\icmlauthor{Drake Thomas}{anthropic}
\icmlauthor{Adrià Garriga-Alonso}{farai}

\end{icmlauthorlist}

\icmlaffiliation{indp}{Independent}
\icmlaffiliation{farlabs}{FAR Labs}
\icmlaffiliation{anthropic}{Anthropic}
\icmlaffiliation{farai}{FAR AI}

\icmlcorrespondingauthor{Thomas Kwa}{kwathomas0@gmail.com}
\icmlcorrespondingauthor{Adrià Garriga-Alonso}{adria@far.ai}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
% \printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

@@

Kullback-Leibler (KL) divergence constraints in reinforcement learning (RL) are employed to stay in regimes where the objective is accurate enough.
Some on-policy \citep{trpo,ppo} and many off-policy \citep{abdolmaleki18mpo,jaques2019way} policy gradient algorithms employ KL constraints or penalties during optimization to prevent the policy from deviating too much from the data collection distribution. This ensures that estimates of each action's advantage are reliable enough to update the policy in a helpful way.

@@

In RLHF, it is common to use the regularization term $\beta D_{KL}\left(\pi \| \pi_0 \right)$ to prevent the learned policy from deviating too much from the base policy, which can prevent unstable behavior or overfitting to the reward model. If our reward model gives reward $U$, then the optimal policy for RLHF with a KL penalty is $
\arg \max_{\pi} \mathbb{E} [U(\pi)]-\beta D_{KL}\left(\pi \| \pi_0 \right).
$

@@

A distribution $P$ over $\mathbb R$ with cumulative distribution function (CDF) $F_P$ is heavy-tailed if its tail function $\bar F_P(x) \triangleq 1 - F_P(x)$ satisfies \(
\lim _{x \rightarrow \infty} e^{t x} \bar{F}(x)=\infty\) for all $t > 0$. 
Heavy-tailed distributions are well-known in statistics to have a higher probability of producing a single extreme value. For example, if the sum of two independent variables from heavy-tailed distributions is large, it is most likely due to one extreme sample rather than two equally large samples. \citep{wierman2013catastrophes}

@@

@@

Reward misspecification is also studied by \citep{lambert2024alignment}, \citep{laidlaw2024preventing}, and others. Laidlaw et al show that a KL penalty between action distributions can be ineffective, and propose instead regularizing state occupancy measure. Our results show an inherent weakness of KL divergence, including when applied to state occupancy measure.

@@

Theorems~\ref{thm5} and \ref{thm6} (Section~\ref{sec:theoretical-conditioning} of the appendix) show that the relationship of catastrophic Goodhart to heavy-tailed error is not just a quirk of KL divergence by using a different model of optimization based on conditioning on high reward values. Under this model (and given additional regularity conditions), it is also true that heavy-tailed error results in catastrophic Goodhart, and light-tailed error plus independence results in arbitrarily large utility. All proofs are in the appendix.

\subsection{Heavy-tailed distributions}
\begin{theorem}
\label{thm1}
Given any heavy-tailed reference distribution
\(Q\) over \(\mathbb R\) with mean \(\mu_Q\), and any
\(M, \epsilon > 0\), there is a distribution \(P\) with mean \(\mu_P>M\)
and \(D_{KL}(P \| Q) < \epsilon\).
\end{theorem}

@@

\hypertarget{theorem-about-RLHF-with-KL-penalty}{%
\subsection{RLHF with KL penalty under heavy-tailed return distribution}}

We now adapt our result to the case where the policy is a language model and we are training it using RLHF. We are now applying KL divergence over the policies rather than the return distributions. We first formally define the properties of RLHF on language models that cause the result to hold: namely, when when considered as a Markov decision process (MDP), environmental transitions are deterministic and return depends only on the final state reached.

@@

\begin{itemize}
\item
  The transition function
  \(P: \mathcal{S} \times \mathcal{A} \to \mathcal{S}\) is
  deterministic, i.e., for each state \(s \in \mathcal{S}\) and action
  \(a \in \mathcal{A}\), there exists a unique state
  \(s' \in \mathcal{S}\) such that \(P(s' | s, a) = 1\). \\ \textbf{In RLHF:} the transition is appending the generated token $a$ to the context $s$.
\item
  There is a set of sink states \(E \subseteq \mathcal S\) that
  terminate a trajectory, which is disjoint from the set of start
  states. \\ \textbf{In RLHF:} The sink states are sequences ending in \texttt{<EOS>} or above a certain length.
\item
  Returns are Markovian; that is, for any two trajectories
  \(\tau=(s_1, a_1, \dots, s_n), \tau'=(s_1', a_1', \dots, s_n'),\) if
  \(s_{n}= s'_{n}\), then \(\tau\) and \(\tau'\) have identical return
  distributions. Equivalently, for the trajectory random variable
  \(T=(S_1, A_1, \dots)\) distributed according to any policy, with
  return \(G\), \(G \bot\!\!\!\!\!\ \bot (S_{<i}, A_{<i}) \ |\ S_i\) for
  any \(i \ge 1\). \\ \textbf{In RLHF:} the return only depends on the full generated string, which is the final state.
\end{itemize}

The language model stochastically outputs the next token $a$ given $s$, and corresponds to the policy.
A DMRMDP is therefore a good model of RLHF.

\begin{theorem}
\label{thm2}
Let \(W = (\mathcal S, \mathcal A, P, R)\) be
a deterministic-transition MDP with Markovian returns. Given \(W\) we
define the function that takes policies to trajectories
\(Tr: (S \to \Delta A) \to \Delta(S \times A)^*\), and the average
return function \(g: (S \times A)^* \to \mathbb R\), which induces a
function \(G: \Delta(S \times A)^* \to \Delta \mathbb R\). Let
\(\pi_0: \mathcal S \to \Delta \mathcal A\) be some base policy. If
\(G \circ Tr(\pi_0)\) is heavy-tailed with finite mean \(\mu_Q\), then
for any \(M, \epsilon > 0\), there is a policy \(\pi\) with mean return
\(\mathbb E[U | U \sim G \circ Tr(\pi)] > M\) and
\(\mathbb E_{s \in T, T\sim Tr(\pi)}[D_{KL}(\pi(s) \| \pi_0(s))] < \epsilon\).
\end{theorem}

\subsection{If V is light-tailed, \texorpdfstring{$\mathbb E_P[V]\! -\! E_Q[V] \to 0$}{} as \texorpdfstring{$D_{KL}\! \to\! 0$}{}}

\begin{theorem}
\label{thm3} If \(V\) is light-tailed and \(d = D_{KL}(P \| Q)\) is bounded, then
\(\mathbb E_P[V]\) is bounded, and \(\mathbb E_P[V] - \mathbb E_Q[V] \to 0\) as
\(d \to 0\).
\end{theorem}

\newtheorem{corr}{Corollary}
\begin{corr}
\label{cor:heavy-tailed}
    Theorems~\ref{thm2} and \ref{thm3} imply that when utility is light-tailed, reward modeling errors make the proxy reward heavy-tailed, and a policy $\pi$ is regularized severely enough to have KL divergence values approaching zero, the reward $\mathbb E[U(\pi)]$ can go to infinity while utility $\mathbb E[V(\pi)]$ approaches a value no higher than the base policy.
\end{corr}

\hypertarget{light-tails-independence-imply-mathbb-ev-to-infty}{%
\subsection{\texorpdfstring{Light-tailed + independence imply
\(\mathbb E[V] \to \infty\)}{Light-tailed + independence imply \textbackslash mathbb EV \textbackslash to \textbackslash infty}}\label{light-tails-independence-imply-mathbb-ev-to-infty}}

@@

@@

To empirically test whether the reward is heavy-tailed, we consider two lines of evidence: examining the distributions directly through random sampling and temperature-1 sampling, and finding adversarial token sequences that get high rewards. We examine one small and one medium reward model that performed reasonably well on RewardBench \citep{lambert2023rewardbench}. The small model is an OpenAssistant model based on Pythia 1.4B, and the medium model is Starling 7B-alpha \citep{starling2023}\footnote{https://huggingface.co/berkeley-nest/Starling-RM-7B-alpha}.

For random sampling, we sample 30000 length-1024 sequences of uniformly random tokens and observe the distribution of rewards assigned by both Pythia 1.4B and Llama 7B-chat. We also use Llama 7B-chat to generate 16000 length-133 sequences and observe the distribution of rewards assigned by Starling 7B-alpha.

Because sampling is inefficient at probing the extreme tail, we also find token sequences that optimize Starling 7B-alpha for reward. We considered Greedy Coordinate Gradient (GCG) from \citep{zou2023universal}, a method used to find adversarial suffixes that circumvent jailbreaking, but decided on a faster version of GCG called Accelerated Coordinate Gradient (ACG) from \citep{haizelabs2024acg}. See Table \ref{table1} for ACG hyperparameters.

@@

When sampling token sequences, both the Pythia model on random inputs (Figure \ref{fig:pythia-random}) and Starling 7B-alpha on Llama-generated inputs (Figure \ref{fig:starling-llama}) appear approximately normal and, therefore, light-tailed. Starling on random inputs (Figure \ref{fig:starling-random} is ambiguous, with the exponential Q-Q plot having an outlier that could indicate a heavy-tailed distribution, but the Hill estimator is consistent with a light-tailed distribution. Because Llama-7B-chat is a more reasonable base model than a completely random policy, we believe that Starling 7B-alpha is more likely to be light-tailed for the purposes of our theoretical results.

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth]{images/reward_plots_random_starling_30k_se.png}
    \caption{Plots of the distribution of reward from 30000 random length-1024 token sequences to Starling 7B-alpha. Clockwise from top left: The histogram shows a unimodal distribution with a slight right skew. The normal probability plot indicates the data are heavier-tailed than normal. The Hill estimator (error bars are standard error) appears to be 0.20 for higher values but fluctuates for lower values. The exponential probability plot of the right half of the distribution is consistent with either light or heavy tails (under heavy tails, the slope would go to infinity).}
    \label{fig:starling-random}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth]{images/reward_plots_llama_starling_16k.png}
    \caption{Plots of the reward distribution from 16000 token sequences generated by Llama 7B-chat of length $\le 133$, starting with five random tokens. Clockwise from top left: A histogram shows the reward distribution has a left skew. The normal probability plot suggests reward is approximately normal and thus light-tailed. The Hill estimator plot should stabilize if the distribution is heavy-tailed, but it does not; thus, there is no evidence the distribution is heavy-tailed. The exponential probability plot also indicates light tails, because the curve is bending downwards. }
    \label{fig:starling-llama}
\end{figure}

The ACG results need some interpretation. The KL divergence between two distributions $P$ and $Q$ if $P$ is the same as $Q$ a fraction $1-\alpha$ of the time, but is some value $x$ a fraction $\alpha$ of the time is given by \(D_{KL}(P \| Q) = \left[(1-\alpha)q(x) + \alpha\right] \log \left(\frac{(1-\alpha)q(x) + \alpha}{q(x)}\right) + 
(1-\alpha) \log (1-\alpha) (1 - q(x))\).

When $\alpha$ is small but much larger than $q(x)$, we approximate this to first order as $D_{KL}(P \| Q) \approx \alpha \log \left(\frac{\alpha}{q(x)}\right)$. In Theorems 1 and 2, we prove that when the error is sufficiently heavy-tailed, a policy that gets extremely large reward a small fraction of the time will achieve high expected reward with low KL divergence. This is not the case here because the rewards achieved through ACG were small and the log-probabilities extremely negative. For example, a policy that matches Llama 2-chat's base reward 99\% of the time and uses the highest-reward input generated by ACG $\alpha=$1\% of the time will have KL divergence from Llama 2-chat of $\alpha(\log(\alpha) - 1339.70) = 13.35$ nats, but reward only about $\alpha*(2.2377-0.3329) = 0.02571$ greater than the base model, far less than can be obtained with the same KL divergence by conditioning.

@@

The low-KL policies that result in catastrophic Goodhart are not a unique optimal policy, just one family of high-performing policies. When optimizing $\mathbb{E} [U(\pi)]-\beta D_{K L}\left(\pi, \pi_0 \right)$, the outcome depends on RL training dynamics; it could be that $D_{KL} \to 0$ causing catastrophic Goodhart, but more likely both terms will go to infinity, potentially allowing $V \to \infty$. Catastrophic Goodhart can be prevented by using a light-tailed or bounded reward function.

@@

Theorems 2 and 3 do not require any independence assumption, but Theorems 4, 5, and 6 require that error $X$ and utility $V$ are independent, which seems to be violated in practice. Future work could weaken this assumption, although intuitively obvious ways to weaken it result in the statement being false. \footnote{Suppose that error $X$ is light-tailed conditional on any value of $V$, but our proxy is merely unbiased ($\mathbb E[X|V=v]=0$ for all $v$). Then the limit of $V$ under optimization for $X+V$ still depends on the relationship between $X$ and $V$. If they are independent, Theorem 6 says that \(\lim_{t \to\infty} \mathbb E[V | X + V \ge t] = \infty\). But if \(V \sim N(0, 1)\), and \(X | V \sim N(0, 4)\) when \(V \in [-1, 1]\), otherwise \(X=0\), then \(\lim_{t \to\infty} \mathbb E[V | X + V \ge t] = 0\).}

@@

We did not search the entire space of token sequences, so we cannot rule out that the reward is heavy-tailed enough to cause catastrophic Goodhart in some situations. While it is intractable to search the more than $10^{2000}$ possible token sequences, future work could get more evidence through more powerful optimization methods.

@@

We have argued that the purpose of the KL divergence regularization in RLHF is to mitigate reward misspecification. However, we have also proven that when errors in the reward function are heavy-tailed, it cannot serve this purpose: even with zero KL divergence, there are policies that achieve very high misspecified reward and no actual reward.
